---
title: "Aligning sequences with genoaligner"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aligning sequences with genoaligner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6)
library(genoaligner)
```

`genoaligner` aligns pairs of sequences. It is a **pairwise** aligner: it
compares two sequences you already have, it does not build a multiple alignment
and it does not recruit candidates.

Two entry points cover the two standard pairwise problems:

* `align()` — **edit distance** (Levenshtein; the WFA-equivalent score) with an
  optional CIGAR.
* `align_sw()` — **Smith-Waterman** local alignment with affine gaps, CIGAR,
  and the aligned span's coordinates.

## Edit distance

```{r}
align("ACGTACGT", "ACGTTCGT", smax = 8)
```

By default `smax = 64`: a pair is *resolved* only when its true equal-cost edit
distance is at most `smax`. If it exceeds the bound the pair comes back
`resolved = FALSE` with an `NA` score rather than a wrong number — size `smax`
to your data to avoid silently under-serving.

Both functions are vectorised over the pair columns, so they map straight into a
pipeline:

```{r}
pairs <- data.frame(
  query     = c("ACGTACGT", "GGGG",    "TTTTCCCC", "AAAACCCC"),
  reference = c("ACGTTCGT", "ACGT",    "TTTTCGCG", "AAAATTTT")
)
pairs$result <- align(pairs$query, pairs$reference, smax = 8)
pairs
```

## Smith-Waterman local alignment

```{r}
align_sw("TTTACGTGTT", "ACGTGT", scoring = c(2, -3, 5, 2))
```

`scoring` is `(match, mismatch, gap_open, gap_extend)`; penalties are positive
and subtracted. The result includes the 0-based coordinates of the aligned span
and the CIGAR over that span.

## The CIGAR convention

Identical for both entry points: `M` and `X` consume a text and a pattern base
(`M` on equal, `X` on differing characters); `I` consumes an extra **text** base;
`D` consumes an extra **pattern** base. Empty inputs and `score == 0`
(no positive-scoring local alignment) are reported with `NA`/`-1` coordinates.

## GPU backend

The workhorse of this package is a self-contained CPU core that needs no GPU
toolchain. The wavefront/GPU backend (ROCm and CUDA) computes the same results
with acceleration and is tracked in the sibling C++ `genoaligner` repository.