---
title: "Codon-aware MSA: hare cytochrome b"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Codon-aware MSA: hare cytochrome b}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6)
library(genoaligner)
```

The `lepus-cytb` vignette used bounded edit distance to check which records
really are cytochrome b. The next step in any marker pipeline is the multiple
sequence alignment — and for a *coding* gene the choice of aligner mode is a
biological decision, not a cosmetic one:

> Frameshift-indifferent alignment inserts indels anywhere, breaking the
> reading frame and turning a clean indel history into apparent nonsense.
> Codon-aware alignment only ever inserts whole codons, so a real in-frame
> deletion looks like what it is.

`msa_codon()` does the second thing. This vignette aligns the 37 real cytb
records from `?lepus_cytb` under the vertebrate mitochondrial code
(`gc = 2` — TGA is tryptophan here, not a stop).

## The alignment

```{r}
data(lepus_cytb)

a <- msa_codon(lepus_cytb$sequence, names = lepus_cytb$accession, gc = 2)
a
```

One call, ~2.5 seconds, a 37 × `r ncol(a$aligned)` alignment. Because every
indel is a whole codon, every row stays a multiple of 3:

```{r}
stopifnot(ncol(a$aligned) %% 3 == 0)
```

## What the QC already knows

`msa_codon` tokenizes each sequence to codons before aligning, and reports
what it found — the chosen reading `frame`, the `n_stops` count in that
frame, and `partial` (a trailing partial codon or held-out frameshift
blocks). The QC travels with the result; no separate pass is needed:

```{r}
qc <- cbind(lepus_cytb[c("accession", "species")], a$qc)
qc[order(-qc$n_stops, -qc$partial), ][1:8, ]
```

Perfect cytb records score `n_stops = 0, partial = 0`. The records that do
not are exactly the ones the pairwise vignette flags as suspects — records
whose best frame still contains a stop codon are typically the mislabelled
or non-cytb accessions, and `partial > 0` marks fragments that do not end on
a codon boundary. That is the same QC story MACSE reports, emitted here as
an ordinary data.frame column.

## Frame integrity: codon mode vs plain DNA mode

Align the same records as plain DNA:

```{r}
d <- msa_align(lepus_cytb$sequence, names = lepus_cytb$accession)
d
```

Both alignments are valid outputs of their own rules — but they answer
different questions. In DNA mode an indel can be any width; in codon mode it
cannot. To see it concretely, take two sequences where one carries a
two-codon deletion:

```{r}
# a complete-cds record vs a fragment missing two codons
full <- lepus_cytb$sequence[2]
frag <- paste0(substr(full, 1, 60), substr(full, 67, nchar(full)))

cod <- as.matrix(msa_codon(c(full = full, frag = frag), gc = 2))
dna <- as.matrix(msa_align(c(full = full, frag = frag)))

# first gap position in each alignment
gap <- function(row) min(which(row == "-"))
cat("codon-mode gap starts at column:", gap(cod["frag", ]), "\n")
cat("  (multiple of 3 ->", (gap(cod["frag", ]) - 1) %% 3 == 0, ")\n")
cat("dna-mode   gap starts at column:", gap(dna["frag", ]), "\n")
```

In codon mode the deletion is placed as codons — the aligned fragment reads
as clean amino acids end to end. That is the difference that matters when
the alignment feeds a partition file, a codon model, or a tree search.

## Frameshift refinement

`msa_codon(..., refine = TRUE)` adds a MACSE-style stage-2 pass that
reintroduces real 1-2 nt frameshift events as narrow columns instead of
forcing them through whole-codon indels. The result is a nucleotide MSA that
is *not* guaranteed to be a multiple of 3 — on purpose, because sequences
that genuinely carry frameshifts should not be aligned as if they did not:

```{r}
r <- msa_codon(c("ATGATAATCACCTGA", "ATGATTACACCTGA"), gc = 2, refine = TRUE)
as.character(r)
```

## Interoperability

The result is an ordinary character matrix underneath:

```{r}
m <- as.matrix(a)
m[1:2, 1:24]
```

`as.character(a)` emits FASTA records ready for `writeLines()`, and with
**ape** installed `as_dnabin(a)` coerces straight to `DNAbin`:

```{r, eval=requireNamespace("ape", quietly = TRUE)}
library(ape)
bin <- as_dnabin(a)
bin[1:3, 1:15]
```

## Where this engine lives

The R package ships the host path of the same C++17 core the GPU library
uses. On the sibling repository the identical engine runs as a level-batched
GPU driver verified bit-exact against this host path — so an alignment
produced here and one produced on the GPU cluster are byte-identical, not
merely similar.
