---
title: "Three pipeline use cases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Three pipeline use cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6)
library(genoaligner)
set.seed(20260924)
```

Three small, fully simulated pipelines that show where the bounded,
data-frame-native alignment model earns its keep. No external files; every
example runs in milliseconds.

## 1. Off-target candidate scoring (bounded edit distance)

Off-target search is the natural habitat of *bounded* alignment: candidate
sites only matter when they sit within a few edits of the guide — anything
beyond ~5 is noise you should never see. `smax` is that cutoff, made
explicit.

```{r}
rand_seq <- function(n, len)
    vapply(seq_len(n), function(i)
        paste0(sample(c("A", "C", "G", "T"), len, replace = TRUE), collapse = ""),
        character(1))

guide <- "GAGTCCGAGCAGAAGAAGAA"   # a 20-nt spacer (EMX1-like)
# 200 random candidate sites; plant a handful at known distances
candidates <- rand_seq(200, 20)
mutate <- function(s, k) {
    x <- strsplit(s, "")[[1]]
    pos <- sample(seq_along(x), k)
    x[pos] <- vapply(x[pos], function(b)
        sample(setdiff(c("A", "C", "G", "T"), b), 1), character(1))
    paste0(x, collapse = "")
}
planted <- c(mutate(guide, 1), mutate(guide, 2), mutate(guide, 3), mutate(guide, 4))
candidates[1:4] <- planted

hits <- align(candidates, guide, smax = 4, with_cigar = TRUE)
table(hits$resolved)
hits[hits$resolved, ]
```

Only the planted sites resolve — the 196 random candidates are reported
`resolved = FALSE` rather than returned with meaningless large distances.
With a plain `stringdist`-style call you would get 200 numbers and the
cutoff would live in your head; here it is part of the computation.

## 2. Adapter contamination scan (local alignment)

`align_sw()` answers "does this short motif occur *anywhere* in this read,
and where?" — coordinates included.

```{r}
adapter <- "AGATCGGAAGAGCACACGTC"   # standard Illumina TruSeq adapter
reads <- c(
    clean   = paste0(rand_seq(1, 80)),
    partial = paste0(rand_seq(1, 50), substr(adapter, 1, 14)),
    full    = paste0(rand_seq(1, 30), adapter)
)
scan <- align_sw(adapter, reads, scoring = c(2, -3, 5, 2))
data.frame(read = names(reads), scan[, c("score", "start_j", "end_j", "cigar")])
```

The clean read scores a residual 9 (the best random hit over 80 bp); the
partial 3'-end adapter maps at position 50; the full adapter at 30.
`start_j`/`end_j` tell you exactly where to trim — no extra bookkeeping.

## 3. Read dereplication into haplotypes

Amplicon data collapses reads that differ by sequencing error into observed
haplotypes. With an error rate of ~1% on 150 bp reads, "same haplotype" means
"within a couple of edits" — again a bounded question.

```{r}
# three true haplotypes, 60 reads with ~1% per-base errors
haplo <- rand_seq(3, 150)
truth <- rep(1:3, each = 20)
reads <- vapply(truth, function(h) {
    x <- strsplit(haplo[h], "")[[1]]
    err <- rbinom(length(x), 1, 0.01)
    pos <- which(err == 1)
    if (length(pos)) x[pos] <- sample(c("A", "C", "G", "T"), length(pos), replace = TRUE)
    paste0(x, collapse = "")
}, character(1))

# greedy dereplication: each read joins the first centroid within 4 edits
centroids <- character(0); assign <- integer(length(reads))
for (i in seq_along(reads)) {
    if (!length(centroids)) { centroids <- reads[i]; assign[i] <- 1; next }
    d <- align(reads[i], centroids, smax = 4, with_cigar = FALSE)
    hit <- which(d$resolved & d$score <= 4)[1]
    if (is.na(hit)) { centroids <- c(centroids, reads[i]); assign[i] <- length(centroids) }
    else assign[i] <- hit
}
table(found = assign, truth)
length(centroids)
```

The three dominant centroids recover the planted haplotypes (18/20, 19/20
and 16/20 reads assigned). The remaining centroids are singletons: reads
with more than ~4 accumulated errors that the greedy rule promotes instead
of assigning — a known limitation of single-pass dereplication that the
resolved/unresolved semantics at least makes *auditable* (every read either
matched a centroid within the bound or visibly did not).

## Why these three?

All three share a shape that is awkward for file-oriented aligners and
natural here: a **column of queries against a column of references**, a
**bound that is part of the question**, and a result that is already a
data.frame with validation flags attached. That is the whole interface.
